---
title: "Bucketing gift levels and rating levels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bucketing gift levels and rating levels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(scipen = 999)  # Disable scientific notation for large numbers
```

fundr includes reference tables for common gift/ask levels and wealth rating levels, plus helpers to bucket numeric values into consistent labels. The bucketing functions support presets for different organization sizes, from small community foundations to large universities.
## Gift Level Bucketing

The `fundr_gift_levels` table contains comprehensive thresholds from $1 to $150M+:

```{r}
library(fundr)

# View all available thresholds
fundr_gift_levels
```

### Using Presets

Different organizations have different gift level needs. Use `gift_levels()` to get a preset appropriate for your shop:
```{r}
# Small shop: annual fund focus ($1 to $100K)
gift_levels("small")

# Medium shop: leadership/major gifts ($1K to $1M)
gift_levels("medium")

# Large shop: principal gifts ($100K to $150M+)
gift_levels("large")
```

### Bucketing Gift Amounts

Use `bucket_gift_level()` to categorize gift amounts:

```{r}
amounts <- c(250, 1000, 5000, 25000, 100000, 1000000, NA)

# Using comprehensive levels (default)
bucket_gift_level(amounts, what = "giving_level")

# Using a preset for smaller gift ranges
bucket_gift_level(amounts, levels = "small", what = "giving_level")

# Get broader bucket labels
bucket_gift_level(amounts, levels = "small", what = "ask_bucket")
```

### Custom Thresholds

You can also select specific thresholds:

```{r}
# Custom selection of thresholds
my_levels <- gift_levels(include = c(0, 0.01, 1000, 5000, 25000, 100000))
my_levels

bucket_gift_level(c(500, 2500, 50000), levels = my_levels)
```

## Rating Level Bucketing

The `fundr_rating_levels` table contains wealth/capacity rating thresholds:

```{r}
# View all available thresholds
fundr_rating_levels
```

### Using Presets

Rating level presets adjust where the bucket boundaries fall (what counts as "Major" vs "Mid-Level"):

```{r}
# Small shop: Major at $25K+, Principal at $250K+
rating_levels("small")

# Large shop: Major at $100K+, Principal at $5M+
rating_levels("large")
```

### Bucketing Ratings

```{r}
ratings <- c(10000, 50000, 250000, 1000000, 10000000)

# Using comprehensive levels
bucket_rating_level(ratings, what = "rating_bucket")

# Using small shop preset (lower thresholds for Major/Principal)
bucket_rating_level(ratings, levels = "small", what = "rating_bucket")

# Using large shop preset
bucket_rating_level(ratings, levels = "large", what = "rating_bucket")
```

### Custom Bucket Assignments

For full control, create custom bucket mappings:

```{r}
# Custom thresholds with custom bucket assignments
my_ratings <- rating_levels(
  include = c(0, 1, 10000, 50000, 250000),
  bucket_map = c(
    "0" = "Unrated",
    "1" = "Annual",
    "10000" = "Mid-Level",
    "50000" = "Major",
    "250000" = "Principal"
  )
)
my_ratings
```

## Working with a Data Frame

These functions work well in both base R and dplyr workflows:

```{r}
df <- data.frame(
  donor_id = 1:6,
  ask_amount = c(0, 500, 5000, 50000, 250000, 1000000),
  capacity_value = c(0, 8000, 25000, 100000, 500000, 5000000)
)

# Using small shop presets
df$gift_level <- bucket_gift_level(df$ask_amount, levels = "small")
df$rating_bucket <- bucket_rating_level(df$capacity_value,
                                         levels = "small",
                                         what = "rating_bucket")
df
```

### With dplyr (Optional)

```{r eval=FALSE}
library(dplyr)

df |>
  mutate(
    gift_level = bucket_gift_level(ask_amount, levels = "small"),
    rating_bucket = bucket_rating_level(capacity_value,
                                         levels = "small",
                                         what = "rating_bucket")
  )
```

## Choosing the Right Preset

| Preset | Gift Levels | Rating Buckets | Best For |
|--------|-------------|----------------|----------|
| `"small"` | $1 to $100K | Major at $25K+, Principal at $250K+ | Community foundations, small nonprofits |
| `"medium"` | $1K to $1M | Major at $50K+, Principal at $1M+ | Mid-size organizations |
| `"large"` | $100K to $150M+ | Major at $100K+, Principal at $5M+ | Universities, hospitals, large institutions |
| `"all"` | Comprehensive | Default buckets | Custom filtering |

## Summary

| Function | Purpose |
|----------|---------|
| `gift_levels()` | Get gift level presets |
| `bucket_gift_level()` | Categorize gift amounts |
| `rating_levels()` | Get rating level presets |
| `bucket_rating_level()` | Categorize wealth ratings |
| `fundr_gift_levels` | Comprehensive gift threshold table |
| `fundr_rating_levels` | Comprehensive rating threshold table |
