---
title: "Data Cleaning with fundr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Cleaning with fundr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Fundraising databases often contain inconsistent data entry. fundr provides tools to normalize phone numbers, ZIP codes, and convert between naming conventions.

```{r}
library(fundr)
```

## Phone Number Normalization

The `normalize_phone()` function handles the many ways phone numbers get entered:

```{r}
messy_phones <- c(
  "(214) 555-1234",
  "972.555.4567",
  "817-555-8901",
  "1-469-555-0123",
  "+1 (214) 555-9876",
  "2145557890",
  "214/555-4321"
)

normalize_phone(messy_phones)
```

### Output Formats

Choose your preferred output format:

```{r}
phone <- "(214) 555-1234"

# Dashed format (default)
normalize_phone(phone, format = "dash")

# Digits only
normalize_phone(phone, format = "digits")

# E.164 international format
normalize_phone(phone, format = "e164")
```

### Handling Extensions

```{r}
phones_with_ext <- c(
  "214-555-1234 x100",
  "972-555-4567 ext. 200",
  "817-555-8901 extension 300"
)

# Drop extensions (default)
normalize_phone(phones_with_ext)

# Keep extensions
normalize_phone(phones_with_ext, keep_extension = TRUE)
```

### Strict vs. Lenient Mode

By default, `normalize_phone()` requires exactly 10 digits. Use `strict = FALSE` to attempt salvaging numbers with extra digits:

```{r}
problematic <- c("12145551234", "0012145551234")

# Strict mode - returns NA for invalid
normalize_phone(problematic, strict = TRUE)

# Lenient mode - takes last 10 digits
normalize_phone(problematic, strict = FALSE)
```

### 7-Digit Numbers

Some databases store local numbers without area codes:

```{r}
local_numbers <- c("555-1234", "5554567")

# By default, 7-digit numbers are invalid
normalize_phone(local_numbers)

# Allow 7-digit numbers
normalize_phone(local_numbers, allow_7 = TRUE)
```

## ZIP Code Normalization

The `normalize_zip()` function standardizes US ZIP codes and ZIP+4:

```{r}
messy_zips <- c(
  "75201",
  "75201-1234",
  "752011234",
  "75201 5678",
  " 75201 "
)

normalize_zip(messy_zips)
```

### Output Formats

```{r}
zip <- "75201-1234"

# 5-digit only (default)
normalize_zip(zip, format = "zip5")

# ZIP+4 when available
normalize_zip(zip, format = "zip9")

# Raw digits
normalize_zip(zip, format = "digits")
```

### Strict vs. Lenient Mode

```{r}
problematic_zips <- c("7520112345", "0075201")

# Strict mode
normalize_zip(problematic_zips, strict = TRUE)

# Lenient mode - salvages valid portions
normalize_zip(problematic_zips, strict = FALSE)
```

## Case Conversion

Convert strings between common naming conventions:

```{r}
# Snake case (database columns)
to_snake_case("FirstName")
to_snake_case("lastName")
to_snake_case("Date of Birth")

# Title case (display labels)
to_title_case("first_name")
to_title_case("date_of_birth")

# Camel case (programming)
to_camel_case("first_name")
to_camel_case("date of birth")
```

### Converting Data Frame Column Names

Use `convert_names()` to transform all column names at once:

```{r}
df <- data.frame(
  `First Name` = "John",
  `Last Name` = "Smith",
  `Date of Birth` = "1980-01-15",
  check.names = FALSE
)

names(df)

# Convert to snake_case
df_snake <- convert_names(df, case = "snake")
names(df_snake)

# Convert to title case
df_title <- convert_names(df_snake, case = "title")
names(df_title)
```

## Working with the Portfolio Dataset

The `fundr_portfolio` dataset intentionally includes messy phone numbers and ZIP codes for practicing cleanup:

```{r}
# Sample of messy phone numbers
sample_phones <- fundr_portfolio$phone_number[!is.na(fundr_portfolio$phone_number)][1:8]
sample_phones

# Clean them
normalize_phone(sample_phones)
```

```{r}
# Sample of messy ZIP codes
sample_zips <- fundr_portfolio$zip[!is.na(fundr_portfolio$zip)][1:8]
sample_zips

# Clean them
normalize_zip(sample_zips)
```

### Cleaning the Full Dataset

Here's how you might clean the entire portfolio using dplyr:

```{r eval=FALSE}
library(dplyr)

portfolio_clean <- fundr_portfolio |>
  mutate(
    phone_clean = normalize_phone(phone_number),
    zip_clean = normalize_zip(zip, format = "zip5")
  )
```

Or with base R:

```{r}
# Create a copy
portfolio_clean <- fundr_portfolio

# Clean phone and ZIP
portfolio_clean$phone_clean <- normalize_phone(portfolio_clean$phone_number)
portfolio_clean$zip_clean <- normalize_zip(portfolio_clean$zip, format = "zip5")

# Check results
head(portfolio_clean[!is.na(portfolio_clean$phone_number),
                     c("phone_number", "phone_clean", "zip", "zip_clean")])
```

## Handling Invalid Values

Both normalization functions return `NA` by default for invalid values. You can customize this:

```{r}
invalid_phones <- c("123", "not-a-phone", "")

# Default - returns NA
normalize_phone(invalid_phones)

# Custom invalid value
normalize_phone(invalid_phones, invalid = "INVALID")
```

## Summary

| Function | Purpose | Key Options |
|----------|---------|-------------|
| `normalize_phone()` | Standardize phone numbers | `format`, `strict`, `allow_7`, `keep_extension` |
| `normalize_zip()` | Standardize ZIP codes | `format`, `strict` |
| `to_snake_case()` | Convert to snake_case | - |
| `to_title_case()` | Convert to Title Case | - |
| `to_camel_case()` | Convert to camelCase | - |
| `convert_names()` | Convert data frame column names | `to` |
