---
title: "Donor Analytics with fundr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Donor Analytics with fundr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```
This vignette demonstrates how to use fundr for common donor analytics tasks, using the included `fundr_portfolio` dataset.

```{r}
library(fundr)
fundr_setup()
```

## The Portfolio Dataset

`fundr_portfolio` contains 10,000 synthetic constituent records:

```{r}
dim(fundr_portfolio)

# Record types
table(fundr_portfolio$record_type)

# Donor vs non-donor
table(!is.na(fundr_portfolio$first_gift_date))
```

## Donor Status Classification

The `donor_status()` function classifies donors based on when they last gave:

- **Active**: Gave in the current fiscal year
- **LYBUNT**: Last Year But Unfortunately Not This (gave last FY, not this FY)
- **SYBUNT**: Some Year But Unfortunately Not This (gave 2 FYs ago)
- **Lapsed**: Gave more than 2 FYs ago
- **Never**: No giving history

```{r}
# Classify based on last gift date
status <- donor_status(fundr_portfolio$last_gift_date)
table(status, useNA = "ifany")
```

### Understanding the Classification

```{r}
# Sample dates to illustrate classification
sample_dates <- as.Date(c(
  "2025-10-15",  # Current FY - Active
  "2024-09-01",  # Last FY - LYBUNT
  "2023-12-15",  # Two FYs ago - SYBUNT
  "2020-06-30",  # Older - Lapsed
  NA             # No gift - Never
))

donor_status(sample_dates)
```

## Giving Analysis

### Gift Level Bucketing

Categorize gifts by amount using `bucket_gift_level()`:

```{r}
# Sample gift amounts
gifts <- c(25, 100, 500, 1000, 5000, 25000, 100000, 1000000)

# Get giving level labels
bucket_gift_level(gifts, what = "giving_level")

# Get broader buckets
bucket_gift_level(gifts, what = "ask_bucket")
```

### Analyzing Portfolio Giving

```{r}
# Bucket the largest gifts
fundr_portfolio$gift_level <- bucket_gift_level(
  fundr_portfolio$largest_gift_amount,
  what = "giving_level"
)

# Distribution (showing non-NA values)
gift_dist <- table(fundr_portfolio$gift_level)
gift_dist[gift_dist > 0]
```

### Giving Summary Statistics

```{r}
# Filter to donors only
donors <- fundr_portfolio[!is.na(fundr_portfolio$first_gift_date), ]

# Summary statistics
cat("Total Donors:", nrow(donors), "\n")
cat("Total Giving: $", format(sum(donors$total_giving, na.rm = TRUE), big.mark = ","), "\n")
cat("Average Gift: $", format(mean(donors$largest_gift_amount, na.rm = TRUE), big.mark = ","), "\n")
cat("Median Gift: $", format(median(donors$largest_gift_amount, na.rm = TRUE), big.mark = ","), "\n")
```

## Prospect Analysis

### Research Ratings

The portfolio includes wealth/capacity ratings:

```{r}
# Rating distribution
rating_dist <- table(fundr_portfolio$research_rating)
rating_dist
```

### Rating Bucketing

Convert ratings to broader categories:

```{r}
# Get rating buckets
fundr_portfolio$rating_bucket <- bucket_rating_level(
  fundr_portfolio$research_rating
)

table(fundr_portfolio$rating_bucket, useNA = "ifany")
```

### Prospect Pipeline

Analyze the prospect pipeline:

```{r}
# Prospects with assigned fundraisers
assigned <- fundr_portfolio[!is.na(fundr_portfolio$fundraiser), ]
cat("Assigned Prospects:", nrow(assigned), "\n\n")

# Pipeline stages
table(assigned$prospect_status)
```

### Fundraiser Portfolios

```{r}
# Prospects per fundraiser
table(fundr_portfolio$fundraiser, useNA = "ifany")
```

## Currency Formatting

Use `format_currency()` and `format_currency_short()` for display:

```{r}
amounts <- c(1234.56, 50000, 1500000, 25000000)

# Full format
format_currency(amounts)

# Short format (K/M/B)
format_currency_short(amounts)
```

### Percentage Formatting

```{r}
rates <- c(0.156, 0.0234, 0.789)
format_pct(rates)
format_pct(rates, digits = 1)
```

## Constituent Age

Calculate ages from dates of birth:

```{r}
# Using calc_age() for full dates
sample_dobs <- as.Date(c("1960-05-15", "1975-12-01", "1990-03-22"))
calc_age(sample_dobs)

# Using calc_age_partial() for varied formats (from portfolio)
sample_dobs_varied <- c("05/15/1960", "December 1975", "1990")
calc_age_partial(sample_dobs_varied)
```

### Age Distribution in Portfolio

```{r}
# Calculate ages
ages <- calc_age_partial(fundr_portfolio$dob)

# Summary
summary(ages)

# Age buckets
age_buckets <- cut(ages,
  breaks = c(0, 30, 40, 50, 60, 70, 80, Inf),
  labels = c("Under 30", "30-39", "40-49", "50-59", "60-69", "70-79", "80+"),
  right = FALSE
)
table(age_buckets, useNA = "ifany")
```

## Giving Tenure

### Total Years Giving

```{r}
# Distribution of giving tenure
table(donors$total_years_giving, useNA = "ifany")
```

### Consecutive Years

```{r}
# Recent donors with consecutive year streaks
recent_donors <- donors[!is.na(donors$consecutive_years_giving), ]
cat("Donors with active streaks:", nrow(recent_donors), "\n\n")

# Streak distribution
table(recent_donors$consecutive_years_giving)
```

## Regional Analysis

The portfolio has geographic weighting toward Dallas/Fort Worth:

```{r}
table(fundr_portfolio$region)
```

### Giving by Region

```{r}
# Average total giving by region
aggregate(total_giving ~ region, data = donors, FUN = mean, na.rm = TRUE)
```

## Segmentation Example

Combine multiple attributes for donor segmentation:

```{r}
# Create a high-value active donor segment
high_value_active <- fundr_portfolio[
  !is.na(fundr_portfolio$donor_status) &
  fundr_portfolio$donor_status == "Active" &
  !is.na(fundr_portfolio$largest_gift_amount) &
  fundr_portfolio$largest_gift_amount >= 10000,
]

cat("High-Value Active Donors:", nrow(high_value_active), "\n")
cat("Total Giving: $", format(sum(high_value_active$total_giving, na.rm = TRUE), big.mark = ","), "\n")
```

## Using dplyr (Optional)

If you prefer tidyverse workflows:

```{r eval=FALSE}
library(dplyr)

fundr_portfolio |>
  filter(!is.na(first_gift_date)) |>
  mutate(
    donor_status = donor_status(last_gift_date),
    gift_level = bucket_gift_level(largest_gift_amount),
    age = calc_age_partial(dob)
  ) |>
  group_by(donor_status) |>
  summarize(
    n = n(),
    total_giving = sum(total_giving, na.rm = TRUE),
    avg_age = mean(age, na.rm = TRUE)
  )
```

## Summary

| Function | Purpose |
|----------|---------|
| `donor_status()` | Classify donors by recency |
| `bucket_gift_level()` | Categorize gift amounts |
| `bucket_rating_level()` | Categorize wealth ratings |
| `calc_age()` | Calculate age from DOB |
| `calc_age_partial()` | Calculate age from partial DOB |
| `format_currency()` | Format as $1,234.56 |
| `format_currency_short()` | Format as $1.2M |
| `format_pct()` | Format as percentage |
| `years_since()` | Years since a date |
| `total_giving_years()` | Count distinct giving years |
| `consecutive_giving_years()` | Current giving streak |
