---
title: "Working with Fiscal Years and Dates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Fiscal Years and Dates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Nonprofits typically operate on fiscal years that don't align with the calendar year. fundr provides comprehensive tools for working with fiscal years, partial dates (common with birth dates), and other date utilities.

```{r}
library(fundr)
```

## Fiscal Year Functions

Most educational institutions and many nonprofits use a July 1 - June 30 fiscal year. fundr defaults to this convention.

### Basic Fiscal Year Conversion

```{r}
# Key dates around fiscal year boundary
dates <- as.Date(c("2024-06-30", "2024-07-01", "2025-01-15", "2025-06-30"))

# Get the fiscal year number
fy_year(dates)

# Get a formatted label
fy_label(dates)

# Get the fiscal quarter
fy_quarter(dates)
```

Note that June 30, 2024 is in FY24, while July 1, 2024 starts FY25.

### Customizing the Fiscal Year Start

If your organization uses a different fiscal year, specify the starting month:

```{r}
date <- as.Date("2024-10-15")

# July start (default)
fy_year(date, fy_start_month = 7)

# October start (federal fiscal year)
fy_year(date, fy_start_month = 10)

# January start (calendar year)
fy_year(date, fy_start_month = 1)
```

### Setting a Global Default

Use `fundr_setup()` to set your organization's fiscal year start globally:

```{r eval=FALSE}
# Set fiscal year to start in October
fundr_setup(fy_start_month = 10)

# Now all fiscal year functions use October
fy_year(as.Date("2024-10-15"))  # Returns 2025
```

### Fiscal Year Labels

`fy_label()` provides human-readable fiscal year labels:

```{r}
dates <- as.Date(c("2024-08-15", "2025-02-20"))

# Default short format (FY25, FY26)
fy_label(dates)

# With custom prefix
fy_label(dates, prefix = "Fiscal ")

# Full year format
fy_label(dates, short = FALSE)
```

### Fiscal Quarters

`fy_quarter()` returns the fiscal quarter (1-4):

```{r}
# Sample dates throughout fiscal year 2025
dates <- as.Date(c(
  "2024-07-15",  # Q1
  "2024-10-15",  # Q2
  "2025-01-15",  # Q3
  "2025-04-15"   # Q4
))

fy_quarter(dates)
```

## Partial Dates

Birth dates in fundraising databases often have incomplete information - sometimes only the year, or year and month. fundr handles these gracefully.

### Parsing Partial Dates

```{r}
# Various date formats from the portfolio
dob_samples <- c(
  "03/15/1985",      # Full date
  "March 1985",      # Month and year
  "1985",            # Year only
  "1985-03-15",      # ISO format
  "03-15-85"         # Two-digit year
)

parse_partial_date(dob_samples)
```

### Date Precision

Determine what level of precision each date has:

```{r}
dob_samples <- c("03/15/1985", "March 1985", "1985", NA)

date_precision(dob_samples)
```

### Calculating Age with Partial Dates

`calc_age_partial()` calculates age from potentially incomplete dates:

```{r}
dob_samples <- c("03/15/1985", "March 1985", "1985")

calc_age_partial(dob_samples)
```

For year-only dates, the function assumes mid-year (July 1) for the calculation.

### Working with Portfolio DOBs

The portfolio dataset includes dates in various formats:

```{r}
# Sample DOBs from the portfolio
sample_dobs <- fundr_portfolio$dob[!is.na(fundr_portfolio$dob)][1:10]
sample_dobs

# Parse them
parsed <- parse_partial_date(sample_dobs)
parsed

# Check precision
date_precision(sample_dobs)
```

## Date Utilities

fundr includes several helpful date utility functions.

### Time Intervals

Calculate time between dates:
```{r}
start <- as.Date("2020-01-15")
end <- as.Date("2024-06-30")

date_interval(start, end, unit = "years")
date_interval(start, end, unit = "months")
date_interval(start, end, unit = "days")
```

### Years Since

Calculate years since a date (useful for giving recency):

```{r}
last_gift_dates <- as.Date(c("2024-01-15", "2022-06-30", "2018-03-01"))

years_since(last_gift_dates)
```

### Recency Bucketing

Categorize dates by how recent they are:

```{r}
dates <- as.Date(c("2025-12-01", "2024-06-15", "2022-01-01", "2018-06-30"))

bucket_recency(dates)
```

### Checking Date Ranges

Test if dates fall within a specified range:

```{r}
dates <- as.Date(c("2024-01-15", "2024-06-30", "2025-01-15"))

# Within last 6 months?
is_within(dates, within = 6, unit = "months")

# Within last 2 years?
is_within(dates, within = 2, unit = "years")
```

### Finding Weekdays

Useful for scheduling and reporting:

```{r}
date <- as.Date("2025-02-28")

# What day is this?
weekday_name(date)

# Find next Monday
next_weekday("Monday", as_of = date)

# Find last Friday
last_weekday("Friday", as_of = date)
```

## Practical Examples

### Fiscal Year Giving Summary

```{r}
# Add fiscal year to giving data
portfolio_fy <- fundr_portfolio
portfolio_fy$last_gift_fy <- fy_year(portfolio_fy$last_gift_date)

# Count donors by fiscal year
table(portfolio_fy$last_gift_fy, useNA = "ifany")
```

### Age Distribution

```{r}
# Calculate ages from DOBs
ages <- calc_age_partial(fundr_portfolio$dob)

# Summary statistics
summary(ages)
```

### Donor Recency Analysis

```{r}
# Categorize donors by recency
recency <- bucket_recency(fundr_portfolio$last_gift_date)
table(recency, useNA = "ifany")
```

## Summary

| Function | Purpose |
|----------|---------|
| `fy_year()` | Convert date to fiscal year number |
| `fy_label()` | Convert date to fiscal year label (e.g., "FY25") |
| `fy_quarter()` | Get fiscal quarter (1-4) |
| `parse_partial_date()` | Parse dates with varying precision |
| `date_precision()` | Determine date precision level |
| `calc_age_partial()` | Calculate age from partial dates |
| `date_interval()` | Calculate time between dates |
| `years_since()` | Years elapsed since a date |
| `bucket_recency()` | Categorize by recency |
| `is_within()` | Check if within time range |
| `last_weekday()` | Find previous occurrence of weekday |
| `next_weekday()` | Find next occurrence of weekday |
| `weekday_name()` | Get day of week name |
