---
title: "Introduction to fundr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fundr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

fundr is an R package providing lightweight, tidyverse-compatible utilities for fundraising and advancement analytics. It's designed with minimal dependencies while offering powerful tools for common nonprofit data tasks.
## Installation

```{r eval=FALSE}
# Install from GitHub
# install.packages("pak")
pak::pak("mattfarrow1/fundr")
```

## Quick Start

Load the package and set up your session:

```{r}
library(fundr)

# Configure session for fundraising work
fundr_setup()
```

`fundr_setup()` disables scientific notation so large dollar amounts display properly (e.g., `$1,000,000` instead of `1e+06`).

## Sample Data

fundr includes `fundr_portfolio`, a synthetic dataset of 10,000 constituent records for learning and testing:
```{r}
dim(fundr_portfolio)
names(fundr_portfolio)[1:15]
```

The dataset includes individuals (singles and married couples) and organizations with realistic data quality variations:

```{r}
table(fundr_portfolio$record_type)
```

## Package Overview

### Data Cleaning

**Phone normalization** - Clean messy phone numbers:

```{r}
messy_phones <- c("(214) 555-1234", "972.555.4567", "1-817-555-8901")
normalize_phone(messy_phones)
```

**ZIP code normalization** - Standardize postal codes:

```{r}
messy_zips <- c("75201-1234", "75201 5678", " 75201 ")
normalize_zip(messy_zips)
```

**Case conversion** - Convert between naming conventions:

```{r}
to_snake_case("FirstName")
to_title_case("first_name")
```

### Fiscal Year Functions

Work with fiscal years (default July start):

```{r}
dates <- as.Date(c("2024-06-30", "2024-07-01", "2025-01-15"))

fy_year(dates)
fy_label(dates)
fy_quarter(dates)
```

### Donor Analytics

Classify donors by recency:

```{r}
last_gifts <- as.Date(c("2025-10-15", "2024-08-01", "2023-05-20", "2020-01-01", NA))
donor_status(last_gifts)
```

Calculate constituent age:

```{r}
calc_age(as.Date("1985-03-15"))
```

### Currency Formatting

Format numbers as currency:

```{r}
amounts <- c(1234.56, 1500000, 25000000)

format_currency(amounts)
format_currency_short(amounts)
```

### Bucketing

Categorize gifts and ratings:

```{r}
gift_amounts <- c(500, 50000, 1000000)
bucket_gift_level(gift_amounts, what = "giving_level")
bucket_gift_level(gift_amounts, what = "ask_bucket")
```

### Visualization (requires ggplot2)
fundr provides a custom theme, color palettes, and currency scales:

```{r eval=FALSE}
library(ggplot2)

ggplot(data, aes(x = year, y = amount)) +
geom_col(fill = fundr_colors("teal")) +
  scale_y_currency(short = TRUE) +
  theme_fundr()
```

## Learning More

Explore the other vignettes for deeper dives:

- `vignette("data-cleaning")` - Phone, ZIP, and case conversion
- `vignette("fiscal-years-dates")` - Working with dates and fiscal years
- `vignette("donor-analytics")` - Analyzing the portfolio dataset
- `vignette("visualization")` - Themes, colors, and scales

## Convenience Operators

fundr includes helpful operators:

```{r}
# "not in" operator (opposite of %in%)
c("a", "b", "c") %notin% c("b", "d")
```

And comment formatting helpers for scripts:

```{r}
comment_header("Data Import")
comment_divider()
```
