---
title: "Prepare for a Portfolio Review"
format: html
editor: visual
---

## Setup

One of the goals of `fundr` is to simply the tasks associated with Prospect Development professionals. As a result, many of the functions are perfect for the work of processing a fundraiser's portfolio for an annual review.

```{r}
# Load libraries
library(tidyverse)
library(fundr)

# Initialize defaults
fundr_setup()
```

## Focusing Your Energy

With so many data points available to us, it can be difficult to know where to focus and prioritize your efforts in a portfolio review, especially if you have limited time. One of the easiest ways to simply large amounts of data is to put it into buckets. The `fundr` package includes three functions to help simplify this work.

- `bucket_gift_level`

- `bucket_rating_level`

- `bucket_recency`

### Bucket Gift Level

```{r}
fundr_portfolio |>
  select(largest_gift_amount) |> 
  mutate(largest_gift_bucket = bucket_gift_level(largest_gift_amount,
                                                 levels = "large")) |> 
  ggplot(aes(largest_gift_bucket)) +
  geom_bar()
```

### Bucket Dates

```{r}
fundr_portfolio |> 
  filter(fundraiser == "James Wilson") |> 
  mutate(time_since_assignment = bucket_recency(fundraiser_start_date)) |> 
  ggplot(aes(time_since_assignment)) +
  geom_bar()
```

### Bucket Ratings

For `fundr_rating_levels` I need to do the following:

- Make sure research ratings match rating levels

- Consider renaming rating level to rating

For `bucket_rating_level`, let's change the default for missing values to "U - Unrated" instead of `NA`.

```{r}
fundr_portfolio |> 
  filter(fundraiser == "James Wilson") |> 
  select(research_rating) |> 
  left_join(fundr_rating_levels, by = c("research_rating" = "rating_level"))
  # mutate(rating_bucket = bucket_rating_level(research_rating, 
  #                                            levels = "large")) |> 
  # ggplot(aes(rating_bucket)) +
  # geom_bar()
```
